---
title: "Semaine 3 : L'Évolution Théologique des Papes"
author: "Cyril"
format: gfm
---

Welcome to my 3rd week TidyTuesday analysis, today we are focusing on papals !

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(patchwork)
library(scales)
library(RColorBrewer)
library(tidytext)

encyclicals <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-06-23/encyclicals.csv')
papal_encyclicals <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-06-23/papal_encyclicals.csv')
scripture_references <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-06-23/scripture_references.csv')
```

Let's start our study with an analysis of encyclicals' content by focusing first on the amount of words and sentences in texts by comparing "Rerum Novarum" (1891) and "Magnifica Humanitas" (2025).

```{r}
average <- encyclicals |> 
  summarize(
    average_number_words = mean(word_count, na.rm = TRUE),
    average_number_sentences = mean(sentence_count, na.rm =TRUE),
    .by = encyclical
  )

g1 <- average |> 
  ggplot(aes(x = fct_reorder(encyclical, average_number_words, .desc = TRUE), y = average_number_words, fill = encyclical)) +
  geom_col() + 
  labs(
    title = "Words",
    x = NULL,
    y = NULL
  ) + 
  guides(fill = "none") +
  scale_fill_manual(
    values = c("Rerum Novarum" = "#e6550d",
               "Magnifica Humanitas" = "#2c7fb8")
  ) + 
  theme(
    panel.background = element_blank(),
    plot.title = element_text(hjust = 0.5),
    axis.ticks = element_blank(),
  )

g2 <- average |> 
  ggplot(aes(x = fct_reorder(encyclical, average_number_sentences, .desc = TRUE), y = average_number_sentences, fill = encyclical)) +
  geom_col() + 
  labs(
    title = "Sentences",
    x = NULL,
    y = NULL
  ) + 
  guides(fill = "none") +
  scale_fill_manual(
    values = c("Rerum Novarum" = "#e6550d",
               "Magnifica Humanitas" = "#2c7fb8")
  ) + 
  theme(
    panel.background = element_blank(),
    plot.title = element_text(hjust = 0.5),
    axis.ticks = element_blank(),
  )

g1 + g2 +
  plot_annotation(
    title = "AI revolution is shortening encyclical texts", 
    subtitle = "Words and sentences counts have both reduced with years",
    theme = theme(
      plot.title = element_text(face = "bold", size = 18),
      plot.subtitle = element_text(size = 13)
    )
  )

```

As we could expect, the recent technologies such as internet and AI is shortening encyclicals' content. That's a good start but it's not enough, now we might wonder if the choice of words is also changing.

```{r}
`%notin%` <- Negate(`%in%`)

my_words = c("life", "church", "god", "human", "person",
             "dignity", "common", "love", "people","means",
             "christian", "peace", "truth", "religion",
             "world", "humanity")

number_words <- encyclicals |> 
  unnest_tokens(word, text) |> 
  anti_join(stop_words) |> 
  filter(word %notin% my_words) |> 
  count(word, encyclical) |> 
  slice_max(n, n = 10, by = encyclical, with_ties = FALSE)

g3 <- number_words |> 
  filter(encyclical == "Rerum Novarum") |> 
  ggplot(aes(x = n, y = fct_reorder(word, n))) +
  geom_col(fill = "#e6550d") +
  labs(
    title = "Industrial Revolution",
    x = NULL,
    y = NULL
  ) +
  theme(
    panel.background = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.ticks = element_blank()
  )

g4 <- number_words |> 
  filter(encyclical == "Magnifica Humanitas") |> 
  ggplot(aes(x = n, y = fct_reorder(word, n))) +
  geom_col(fill = "#2c7fb8") +
  labs(
    title = "AI Revolution",
    x = NULL,
    y = NULL
  ) +
  theme(
    panel.background = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.ticks = element_blank()
  )

g3 + g4 +
  plot_annotation(
    title = "Drastic priorities changes in encyclicals",
    subtitle = "Comparison of words occurences in Rerum Novarum and Magnifica Humanitas",
    theme = theme(
      plot.title = element_text(face = "bold", size = 18),
      plot.subtitle = element_text(size = 13)
    )
  )
```

As we can see, the new AI Revolution is shifting priorities of papal's encyclicals content focusing less on workers and more on geopolitics matters, with the disappearence of words like "labor" and "classes" and the uprising of words like "economic" and "digital".

What about which books Leo XIII (Rerum Novarum) and Leo XIV (Magnifica Humanitas) are drawing upon ?

```{r}
books <- scripture_references |> 
  count(
    book,
    pope,
    testament
  ) |> 
  slice_max(n, n = 5, by = pope, with_ties = FALSE) 

g5 <- books |>  
  filter(pope == "Leo XIII") |> 
  ggplot(aes(x = n, y = fct_reorder(book, n), fill = testament)) +
  geom_col() + 
  guides(fill = "none") +
  labs(
    title = "Leo XIII",
    x = NULL,
    y = NULL
  ) +
  scale_fill_manual(
    values = c("Old Testament" = "#d95f02", "New Testament" = "#1b9e77")
  ) +
  theme(
    panel.background = element_blank(),
    axis.ticks = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

g6 <- books |>  
  filter(pope == "Leo XIV") |> 
  ggplot(aes(x = n, y = fct_reorder(book, n), fill = testament)) +
  geom_col() +
  labs(
    title = "Leo XIV",
    x = NULL,
    y = NULL
  ) +
  scale_fill_manual(
    values = c("Old Testament" = "#d95f02", "New Testament" = "#1b9e77")
  ) +
  theme(
    panel.background = element_blank(),
    axis.ticks = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.title = element_blank()
  )

g5 + g6 +
  plot_annotation(
    title = "Nehemiah is becoming important in the AI Revolution",
    subtitle = "Books of the Bible Leo XIII and Leo XIV are drawing upon",
    theme = theme(
      plot.title = element_text(face = "bold", size = 18),
      plot.subtitle = element_text(size = 13)
    )
  )
```

Both papal maintains their baseline of social justice by drawing upon"Matthew", the shift is with the upcoming of Nehemiah that might signal an era that requires protecting the "walls" of human society against unprecedented modern challenges.
